---
title: "Building a Simple Neural Network from Scratch in R"
author: "Your Name"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

## 🧠 Introduction

In this tutorial, we’ll build a **simple neural network from scratch in R** — no deep learning libraries, only base R.  
The goal is to learn the **XOR function**, a classic example that requires a nonlinear model to solve.

| x1 | x2 | y |
|----|----|---|
| 0  | 0  | 0 |
| 0  | 1  | 1 |
| 1  | 0  | 1 |
| 1  | 1  | 0 |

---

## ⚙️ Setup and Activation Functions

We'll define the **sigmoid** activation function and its derivative for backpropagation.

```{r}
# Sigmoid activation and its derivative
sigmoid <- function(x) {
  1 / (1 + exp(-x))
}

sigmoid_derivative <- function(x) {
  x * (1 - x)
}
```


## 📘 Training Data

We’ll use the XOR truth table as input and output matrices.
```{r}

# XOR input and output
X <- matrix(c(0,0,
              0,1), nrow = 2, byrow = TRUE)

y <- matrix(c(0, 1), nrow = 2)

print(X)
```


🧩 Initialize the Network

We define the architecture, hyperparameters, and initialize weights randomly.
```{r}

set.seed(50)

input_neurons <- 2
hidden_neurons <- 2
output_neurons <- 1
learning_rate <- 0.5

# Initialize weights and biases
W1 <- matrix(runif(input_neurons * hidden_neurons), nrow = input_neurons) ## between -1 and 1 
b1 <- matrix(runif(hidden_neurons), nrow = 1)
W2 <- matrix(runif(hidden_neurons * output_neurons), nrow = hidden_neurons)
b2 <- matrix(runif(output_neurons), nrow = 1)

print(W1)
print(b1)
print(W2)
print(b2)

``` 

🔁 Training the Network

We train using forward propagation, error computation, and backpropagation.
```{r}

epochs <- 10000
losses <- numeric(epochs)

for (i in 1:epochs) {
  
  # Forward pass
  z1 <- X %*% W1 + matrix(rep(b1, each = nrow(X)), nrow = nrow(X))
  a1 <- sigmoid(z1)
  z2 <- a1 %*% W2 + matrix(rep(b2, each = nrow(a1)), nrow = nrow(a1))
  a2 <- sigmoid(z2)
  
  # Compute error
  error <- y - a2
  losses[i] <- mean(abs(error))
  
  # Backpropagation
  d_output <- error * sigmoid_derivative(a2)
  d_hidden <- (d_output %*% t(W2)) * sigmoid_derivative(a1)
  
  # Update weights and biases
  W2 <- W2 + t(a1) %*% d_output * learning_rate
  b2 <- b2 + colSums(d_output) * learning_rate
  W1 <- W1 + t(X) %*% d_hidden * learning_rate
  b1 <- b1 + colSums(d_hidden) * learning_rate
  
  # Print loss occasionally
  if (i==1) cat("Initial","\n")

  if (i==1) cat("X0      ",sprintf("%+.2f",X),"\n")


  if (i<3 ) cat("W1      ",sprintf("%+.2f", W1),"\n")

  if (i<3 ) cat("X %*% W1", sprintf("%+.2f",X %*% W1),"\n")
  if (i<3 ) 
for (j in 1:ncol(W1)) {
  for (i in 1:nrow(X)) {
    terms <- paste(paste0(round(X[i, ],2), "*", round(W1[, j],2)), collapse = "+")
    cat(" ", terms)
  }
}
    cat("\n" )


  if (i<3) cat("b1      " , matrix(rep(sprintf("%+.2f",b1), each = nrow(X)), nrow = nrow(X)),"\n")
  if (i<3) cat("z1      ",sprintf("%+.2f",z1),"\n")
  if (i<3) cat("a1      ",sprintf("%+.2f",a1),"\n")

  if (i<3) cat("W2      ",sprintf("%+.2f", W2),"\n")
  if (i<3 ) cat("a1 %*%W2", sprintf("%+.2f",a1 %*% W2),"\n")
  if (i<3 ) 
for (j in 1:ncol(W2))  {
  for (i in 1:nrow(a1)) {
    terms <- paste(paste0(round(a1[i, ],2), "*", round(W2[, j],2)), collapse = "+")
    cat(" ", terms)
  }
}
    cat("\n" )


  if (i<3) cat("b2      " , matrix(rep(sprintf("%+.2f",b2), each = nrow(X)), nrow = nrow(X)),"\n")
  if (i<3) cat("z2      ",sprintf("%+.2f",z2),"\n")

  if (i<3) cat("a2      ",sprintf("%+.2f",a2),"\n")
  
  }
``` 

```{r}

``` 
📉 Visualizing the Training Loss

We can plot the loss curve to see the model improving over time.
```{r}

plot( losses[1:10000], type = "l", col = "blue", lwd = 2,
     xlab = "Epoch", ylab = "Mean Absolute Error",
     main = "Training Loss over Time",     log = "x")

grid()
``` 
✅ Results

After training, let’s check the predictions versus expected outputs.
```{r}

cat("\nFinal Predictions:\n")
print(round(a2, 3))
cat("\nExpected Output:\n")
print(y)

```
You should see predictions close to [0, 1, 1, 0], showing that the network successfully learned XOR.


We’re multiplying two matrices:

\[
X =
\begin{bmatrix}
0 & 0 \\
0 & 1
\end{bmatrix},
\quad
W =
\begin{bmatrix}
1 \\
1
\end{bmatrix}
\]

---

## ⚙️ Matrix Multiplication

We can compute the matrix product:

\[
X \%*\% W =
\begin{bmatrix}
0*1 + 0*1 \\
0*1 + 1*1
\end{bmatrix}
=
\begin{bmatrix}
0 \\
1
\end{bmatrix}
\]

---